Practical Optional Types for Clojure

### 2015-02-26 conclusion.tex
tweak primer. more formal model

We have presented Typed Clojure, an optionally-typed version of
Clojure whose type system works with a wide variety of distinctive
Clojure idioms and features. Although based on the foundation of Typed
Racket's occurrence typing approach, Typed Clojure both extends the
fundamental control-flow based reasoning as well as applying it to
handle seemingly unrelated features such as multi-methods. In
addition, Typed Clojure supports crucial features such as heterogenous
maps and Java interoperability while integrating these features into
the core type system.

>>>>

We have presented Typed Clojure, an optionally-typed version of
Clojure whose type system works with a wide variety of distinctive
Clojure idioms and features. Although based on the foundation of Typed
Racket's occurrence typing approach, Typed Clojure both extends the
fundamental control-flow based reasoning as well as applying it to
handle seemingly unrelated features such as multi-methods. In
addition, Typed Clojure supports crucial features such as heterogeneous
maps and Java interoperability while integrating these features into
the core type system.

### 2015-02-26 formal-model.tex
tweak primer. more formal model

\subsection{Heterogenous maps}
\label{sec:hmapformal}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\BAssoc}
    {\BGet}
    {\BGetMissing}
  \end{mathpar}
  \caption{HMap Semantics}
  \label{main:figure:hmapsem}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap subtyping rules}
  \label{main:figure:hmapsubtype}
\end{figure}

We now present \HMapliteral{} types in detail.
Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
$$
\clj{(HMap :mandatory p :absent-keys a :complete? b)}
$$
is written \HMapgeneric{\mandatory{}}{\absent{}},
where the metavariable \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} otherwise {\partial{}}. 
\mandatory{} specifies a set of known present keyword entries. \absent{} tracks known absent keys.
\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. Keys cannot be both present and absent.

\clj{(get m :a)} is identical to \clj{(:a m)}.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

@@

We introduce our first path element {\keype{\k{}}}, the symbolic representation
of \clj{(get m k)} where \clj{m} represents the path after {\keype{\k{}}}---{\path{\keype{\makekw{b}}}{\path{\keype{\makekw{a}}}{\x{}}}}
  represents \clj{(get (get x :a) :b)}.

Figure~\ref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
T-Assoc extends a HMap with a mandatory entry.
Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
and \assocliteral{}.

S-HMap (figure~\ref{main:figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
subtypes and also cover the absent keys.
Complete maps are subtypes of partial maps by S-HMapP.

@@

Our second path element \classpe{} is the symbolic representation
of a call to \classconst{}---{\path{\classpe{}}{\path{\keype{\makekw{a}}}{\x{}}}}
represents \clj{(class (get x :a))}.
Figure~\ref{main:figure:primitivesem} contains the semantics for \classconst{}.

@@

\subsection{Proof system}

The essential piece of occurrence typing's logical system is L-Update,
where we combine propositions that have similar objects to learn new facts.

\begin{mathpar}
  {\LUpdate}
\end{mathpar}
For example, if \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
and
\inpropenv {\propenv{}} {\notprop {\Nil{}} {\x{}}},
then
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype{\Nil{}}} {\emptypath{}}} {\x{}}},
$$
so by the definition of \updateliteral{} (figure~\ref{main:figure:update})
\inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).

Now that we have our path elements \classpe{} and \keype{\k{}}
and the typing rules that introduce them, we give a detailed example how
each path element interacts with L-Update.

\smallsection{\classpe{}}
Consider the then branch of the following program:

\begin{minted}{clojure}
(fn [x :- (U nil Keyword)]
  (if (class x)
    ; here
    nil))
\end{minted}

Intuitively, since \clj{(class x)} is a true value it must be a class, therefore
\clj{x} is an object (not \clj{nil}).

From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
The test has the object {\path{\classpe{}}{\x{}}} by T-App.
By T-If we can assume
{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
{\path{\classpe{}}{\x{}}}, 
by L-Update 
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
$$
This hits the sixth \classpe{} case of \updateliteral{}, which returns
$$
{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
$$
deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).

\smallsection{\keype{\k{}}}
Consider the then branch of

\begin{minted}{clojure}
(fn [x :- '{:kw (U nil Keyword)}]
  (if (get x :kw)
    ; here
    nil))
\end{minted}

Similar to before,
by T-Abs
{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\x{}}},
the test's object is {\path{\keype{\makekw{kw}}}{\x{}}} by T-App
and by T-If we assume
{\notprop {\falsy{}} {\path{\keype{\makekw{kw}}}{\x{}}}} in the then branch.

L-Update combines these propositions
$$
\inpropenv {\propenv{}} {\isprop {\update {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\nottype {\falsy{}}} {\keype{\makekw{kw}}}} {\x{}}}.
$$
matching the \updateliteral{} case which pushes the \updateliteral{} inside the map
$$
{\HMapp{\mandatorysetsingleton{\makekw{kw}}{\update{\Union{\Nil}{\Keyword}}{\nottype {\falsy{}}}{\emptypath{}}}}{\emptyabsent}}
$$
concluding with
$$
\inpropenv {\propenv{}}
{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Keyword}}{\emptyabsent}}   {\x{}}}.
$$

>>>>

\subsection{Heterogeneous maps}
\label{sec:hmapformal}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\BAssoc}
    {\BGet}
    {\BGetMissing}
  \end{mathpar}
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap Semantics and Subtyping rules}
  \label{main:figure:hmapsem}
\end{figure}

We now present \HMapliteral{} types in detail.
Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
The syntax 
$$
\clj{(HMap :mandatory p :absent-keys a :complete? b)}
$$
we introduced in section~\ref{sec:overview}
is now written \HMapgeneric{\mandatory{}}{\absent{}},
where \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} and is {\partial{}} otherwise. 
The \mandatory{} entry specifies a set of known present keyword entries and \absent{} tracks known absent keys.
To ease presentation, 
if a HMap is declared as \complete{} then \absent{} implicitly contains all other keywords not specified in \mandatory{}.
%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
Keys cannot be both present and absent.

The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
we only model the former to avoid the added complexity of keywords being functions.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

Subtyping for HMaps (figure~\ref{main:figure:hmapsem})
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that a HMap is a subtype of another HMap if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

Figure~\ref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
We only handle unions of HMaps in T-GetHMap, where every element in the union
contains the key we are looking up, in particular to support operations like
looking up \clj{:op} on a map of type \clj{Expr} (example~\ref{example:decleaf}).
The object in each T-Get rule is now more complicated than in T-Local---the 
next section discusses this in detail.
T-Assoc extends a HMap with a mandatory entry.

Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
and \assocliteral{}.
The B-Get rule evaluates first evaluates the map and the key, asserts
the key must be present in the map and returns the associated key. If
the entry is missing, B-GetMissing says we return \nil{}.

@@

Recall the first insight of occurrence typing---not
only do we want to reason about components of the runtime environment, that
is, variables, but \emph{subcomponents} of the environment.
We have already mentioned paths \path{\pathelem{}}{\x{}}
where \pathelem{} is a sequence of selectors on \x{} called \emph{path elements},
but we have not yet defined what components of a path can be.
\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
to reason about selector operations on cons cells.
We instead want to reason about HMap lookups.

We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(figure~\ref{main:figure:hmapsyntax}) by
demonstrating how the first conditional test is checked in example~\ref{example:decleaf}.
First we unfold the destructuring syntax.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (let [op (:op m)]
      (if (= :if (:op m))
        {:op :if, ...}
        (if ...))))
\end{minted}

@@

@@

### 2015-02-26 intro.tex
tweak primer. more formal model

In the remainder of this paper, we describe how Typed Clojure's
central innovations, including Java interoperability, multimethods,
and heterogenously-typed immutable maps, enable this example and many
others. We begin with an example-driven presentation of the main type
system features in Section~\ref{sec:overview}. We then incrementatlly
present a core calculus for Typed Clojure covering all of these
features together in Section~\ref{sec:formal} and prove type
soundness(section \ref{sec:sound}). We then discuss the full
implementation of Typed Clojure, dubbed \coretyped{}, which extends
the formal model in many ways, and the experience gained from its use
in Section~\ref{sec:exp}. Finally, we discuss related work and
conclude.

>>>>

In the remainder of this paper, we describe how Typed Clojure's
central innovations, including Java interoperability, multimethods,
and heterogeneously-typed immutable maps, enable this example and many
others. We begin with an example-driven presentation of the main type
system features in Section~\ref{sec:overview}. We then incrementatlly
present a core calculus for Typed Clojure covering all of these
features together in Section~\ref{sec:formal} and prove type
soundness(section \ref{sec:sound}). We then discuss the full
implementation of Typed Clojure, dubbed \coretyped{}, which extends
the formal model in many ways, and the experience gained from its use
in Section~\ref{sec:exp}. Finally, we discuss related work and
conclude.

### 2015-02-26 overview.tex
tweak primer. more formal model

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if all its arguments are structurally equal---in this
paper we treat it as pointer equality.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

@@

\emph{Commas} are always whitespace.
\emph{Destructuring} is binding position syntax for pattern matching.
A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
to lookups on that keyword with the current argument, optionally terminated by
\clj{:as} and a symbol, which binds the symbol to the current argument.
$$
\clj{(fn [{x :x, y :y, :as m}] ...)}
$$ 
binds \clj{m} to the first argument,
\clj{x} to \clj{(get m :x)} and
\clj{y} to \clj{(get m :y)} in the function body. Type hints are preserved as expected.

@@

Java methods and fields are called in prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.
Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}.

The \clj{if} special form is a standard two-armed conditional.
The \clj{do} special form sequentially evaluates any number of expressions
left-to-right, returning \nil{} on zero arguments.
The \clj{when} macro is a one-armed \clj{if} that returns \clj{nil} if the test fails.
The \clj{throw} special form is exactly Java's \java{throw}, which takes an
instance of \clj{Throwable} and throws an exception with it.
The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
and any number of expressions in an implicit \clj{do} for the body of the function.
The typed
variant of \clj{fn} supports an optional keyword \clj{:-} and type following
each parameter to declare their static types, and similarly after
the parameter vector to declare the return type.

@@

The \clj{defalias} macro defines a type abbreviation which can be recursively defined.

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
\end{exmp}
\end{Code}

>>>>

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if all its arguments are structurally equal---in this
paper we use it only for pointer equality.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

@@

\emph{Commas} are always whitespace.

@@

Java methods and fields are called in prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.
The canonical form for methods is \clj{(. target (method args*))} 
and for fields \clj{(. target field)} using the dot special form, which we utilise occasionally
for clarity.
Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}
or equivalently \clj{(new class args*)} with the \clj{new} special form.

The \clj{if} special form is a standard two-armed conditional.
The \clj{do} special form sequentially evaluates any number of expressions
left-to-right, returning the last expression or \nil{}.
The \clj{when} macro is a one-armed \clj{if} with an implicit \clj{do}
around arguments after the test, and returns \clj{nil} if the test fails.
The \clj{throw} special form is exactly Java's \java{throw}, which takes an
instance of \clj{Throwable} and throws it as an exception.
The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
and any number of expressions in an implicit \clj{do} for the body of the function.
The \clj{let} macro takes a vector of bindings, a flat vector of binding-value pairs,
and scopes them in the body, an implicit \clj{do}.
Scoping works like \clj{let*} in Scheme---that is values in the binding vector can refer
to variables previously introduced in the same binder.
The typed
variant of \clj{fn} supports an optional keyword \clj{:-} and type following
each parameter to declare their static types, and similarly after
the parameter vector to declare the return type.

\emph{Destructuring} is binding position syntax for pattern matching.
A \emph{map destructuring} contains pairs of symbols and keywords that bind the symbols
to lookups on that keyword with the current argument, optionally terminated by
\clj{:as} and a symbol, which binds the symbol to the current argument. For example
$$
\clj{(fn [{^File x :x, ^String y :y, :as m}] ...)}
$$ 
expands to
\begin{minted}{clojure}
  (fn [m___0]
    (let [m m___0,
          ^File x (:x m),
          ^String y (:y m)]
      ...))
\end{minted}
Destructuring is supported similarly by \clj{let}.

@@

The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.

@@

\begin{Code}
\begin{exmp}
\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
\label{example:decleaf}
\end{exmp}
\end{Code}

### 2015-02-26 small.tex
2.5
tweak primer. more formal model

\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
\newcommand{\rkt}[1]{\mintinline{racket}{#1}}

@@

In this paper, we describe Typed Clojure and present these type system
extensions, focusing on three features widely used in Clojure. First, Java
interoperability is central to Clojure's mission but introduces
challenges such as ubiquitous \texttt{null}; Typed Clojure handles
Java interoperability while ensuring the absence of null-pointer
exceptions in typed programs. Second, Clojure programmers
idiomatically use immutable dictionaries for data structures; Typed
Clojure handles this in the type system with multiple forms of
heterogenous dictionary types. Third, multimethods provide extensible
operations, and their Clojure semantics turns out to have a surprising
synergy with the underlying occurrence typing framework.

>>>>

\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
\newcommand{\java}[1]{\mintinline{java}{#1}}
\newcommand{\rkt}[1]{\mintinline{racket}{#1}}

@@

In this paper, we describe Typed Clojure and present these type system
extensions, focusing on three features widely used in Clojure. First, Java
interoperability is central to Clojure's mission but introduces
challenges such as ubiquitous \texttt{null}; Typed Clojure handles
Java interoperability while ensuring the absence of null-pointer
exceptions in typed programs. Second, Clojure programmers
idiomatically use immutable dictionaries for data structures; Typed
Clojure handles this in the type system with multiple forms of
heterogeneous dictionary types. Third, multimethods provide extensible
operations, and their Clojure semantics turns out to have a surprising
synergy with the underlying occurrence typing framework.
